Drought-sensitive trees survive on deep-water access in a tropical forest community

### 2020-08-09 manuscript.rmd
1. From tiff back to jpeg 2. Added SI figure for dbh model 3. Revised description of growth data processing

\begin{equation}
A = a_3WSG^2 + a_2B + a_1B^2 + a_0 + error
(\#eq:a-mod)
\end{equation}

\begin{equation}
B = b_3WSG^2 + b_2WSG.LMA + b_1LMA^2 + b_0 + error
(\#eq:b-mod)
\end{equation}

where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients. $WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. S\@ref(fig:ab-fit)). $WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-vuln-curves)).

@@

## Model for Effective Rooting Depths (ERD)
\
Water is a key limiting resource for tree growth. Vapour pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_leaf$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M4)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows (add additive models xx):

@@

To minimise the effect of light variation among trees, we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) that are likely to be fully exposed to the sun. We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). For each individual tree (i), we calculated growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (xx Add SI figure showing the model). We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx). 

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Fig. (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

### Future Directions for Model Development
\
The trade-offs that we identify between above- and below-ground traits, water environment vertically belowground and mortality are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesised although community-wide examples are rare [@Hacke:2006; @Gleason:2016]. We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species). We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaf, but not branch (add SI figures). High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species). These relationships between ERD and aboveground hydraulic traits will provide an important insight on how to model the rooting depth and its coordination or trade-off with other traits to better represent the functional diversity in tropical forests.

@@

```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017. Red line shows a fit from a polynomial model described in inset, and used to standardise VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
```

@@

```{r obs-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-vuln-curves} Leaf 'vulnerability curves', that is standardized $K_{leaf}$ vs. $\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ ($gcm^{-3}$).")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
```

```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs. $\\Psi_{leaf}$ for ", sp.soft.filled$AB.data.model," species vs. those that were predicted from models based on species soft-traits, WSG and LMA.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.tiff"))
```

@@

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) shown in terms of loss of conductivity (%) (A & B). Each line represents a species, color coded by $WSG$ (A), or $LMA$ (B), with blue tones indicating smaller values and red tones indicating larger values of $WSG$ (A), or $LMA$ (B)")}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg", scale = 1)
cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
```

>>>>

\begin{equation}
A = `r acf[1]` `r acf[2]`WSG `r acf[3]`WSG^2 + `r acf[4]`WSGB + `r acf[5]`B `r acf.6`B^2 + error
(\#eq:a-mod)
\end{equation}

\begin{equation}
B = `r bcf[1]` `r bcf[2]`B + `r bcf[3]`B^2 + `r bcf[4]`BLMA + `r bcf[5]`LMA `r bcf[6]`LMA^2 + error
(\#eq:b-mod)
\end{equation}

$WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. S\@ref(fig:ab-fit)). $WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-vuln-curves)).

@@

## Model for Effective Rooting Depths (ERD)
\
Water is a key limiting resource for tree growth. Vapour pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_leaf$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M4)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows:

@@

For growth estiamtes, to minimise the effect of light variation among trees, we selected only large trees ($\ge$ 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species) and thus are likely to be fully exposed to the sun. We selected trees for which height of diameter measurement was unchanged across all censuses (1990, 1995, 2000, 2005, 2010, 2015). For each individual tree (*i*), we calculated growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. We eliminated positive outliers as any growth rates amounting to > 75 mm.yr^-1^, as this was the highest rate observed with confidence for fastest growing species [@Condit.2017]. Negative outliers were removed based on known measurement error, $\sigma_{e} = 0.006214d + 0.903$, where *d* is dbh. Any stem whose later dbh measurement fell < 4$\sigma_{e}$ the earlier measurement was removed [@Condit.2017]. As tree growth rates vary with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(d_{i, t})$). The diameter model of growth ($f(d_{i, t})$), a B-spline based polynomial equation with 5 degrees of freedom, was fitted to growth rates from all trees pooled together as there was insufficient species-specific data for large trees (Fig. S\@ref(fig:dbh-model))). We retained only those trees that had $Resid_{i,t}$ estimates across all censuses (complete records, total trees = `r sum(g.n$n)`) and standardized (centered and scaled) individual tree time series of $Resid_{i,t}$. We then obtained species-level (*n* = `r length(erd.sp)`) time-series of medians of $Resid_{i,t}$ ($Growth_{t}$). These were obtained for only those species with complete records on at least three trees (median *n* = `r median(g.n$n)`, max *n* = `r max(g.n$n)` trees).

@@

\
Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species; Fig. (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

### Future Directions for Model Development
\
The trade-offs that we identify between above- and below-ground traits, water environment vertically belowground and mortality are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesised although community-wide examples are rare [@Hacke:2006; @Gleason:2016]. We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species). We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaf, but not branch (add SI figures). High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species; xx add figure). These relationships between ERD and aboveground hydraulic traits will provide an important insight on how to model the rooting depth and its coordination or trade-off with other traits to better represent the functional diversity in tropical forests.

@@

```{r gpp-vpd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-gpp-vpd} Scatterplot of mean daily gross primary productivity (GPP) and mean daily vapour pressure deficit (VPD) by month measured at the eddy flux tower in Barro Colorado Island over 2012-2017. Red line shows a fit from a polynomial model described in inset, and used to standardise VPD employed in equations \\@ref(eq:M2) and \\@ref(eq:M4)."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/gpp_by_env_variables/monthly_gpp.vpd.jpeg"))
```

@@

```{r obs-vuln-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-vuln-curves} Leaf 'vulnerability curves', that is, $K_{leaf}$ vs. $\\Psi_{leaf}$, fitted to observed data on ",  length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))," tree species from Panama, color coded by Wood Specific Gravity~100^$\\circ$^C~ (gcm^-3^).")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB_for_data_sp.jpeg"))
```

```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs. $\\Psi_{leaf}$ for ", sp.soft.filled$AB.data.model," species vs. those that were predicted from models based on species soft-traits, WSG and LMA.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
```

```{r dbh-model, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-dbh-model} A scatterplot of tree growth rates (circles) from all census intervals versus dbh at the beginning of the cenus interval is overlaid with a dbh model of growth (spline). Growth rates are color coded by census interval. The model equation is shown in inset along with goodness-of-fit(*R*^2^)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/dbh/growthrate_dbh_predicted_growth_with_spline_degree_5_dbh_cutoff_100_model.jpeg"))
```

@@

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled 'vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) shown in terms of loss of conductivity (%) (A & B). Each line represents a species, color coded by $WSG$ (A), or $LMA$ (B), with blue tones indicating smaller values and red tones indicating larger values of $WSG$ (A), or $LMA$ (B)")}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/std.k.spkmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg", scale = 1)
cowplot::plot_grid(p1, p2, ncol = 2, labels = "AUTO")
```
